Nitrogen's form determines its fate in the North Pacific Subtropical Gyre during PARAGON I

### 2025-05-17 manuscript/manuscript.Rmd
Adding kmeans clustering code and figures

```{r get counts per amendment instead of as a whole}
# Repeat analysis across amendments to get additional numbers
lmtest_xamend <- cleaned_combined_peaks %>%
  arrange(feature) %>%
  select(feature, filename, into) %>%
  mutate(samp_type=str_extract(filename, "Smp|Blk|Poo")) %>%
  filter(samp_type=="Smp") %>%
  mutate(timepoint=as.numeric(str_extract(filename, "(?<=T)\\d+"))) %>%
  mutate(timepoint=ifelse(timepoint==0, 0.3, timepoint)) %>%
  mutate(amendment=str_extract(filename, "Amm|GMP|Arg|Nit")) %>%
  mutate(depth=str_extract(filename, "Surf|Deep")) %>%
  nest(data=-c(feature, amendment)) %>%
  filter(sapply(data, nrow)>2) %>%
  mutate(lm_test=map(data, ~broom::tidy(lm(log10(.x$into)~log10(.x$timepoint))), .progress = TRUE)) %>%
  unnest(lm_test) %>%
  filter(term=="log10(.x$timepoint)") %>%
  mutate(p_adj=p.adjust(p.value, method = "fdr"))
feat_shape_df_xamend <- cleaned_combined_peaks %>%
  mutate(amendment=str_extract(filename, "Amm|GMP|Arg|Nit")) %>%
  summarise(mzmed=unique(mzmed),
            rtmed=unique(rtmed),
            med_iso_cor=median(iso_cor), 
            med_snr=median(beta_snr),
            med_cor=median(beta_cor),
            .by=c(feature, amendment)) %>%
  left_join(lmtest_xamend %>% select(feature, amendment, estimate, p_adj)) %>%
  mutate(peak_qual=ifelse(med_snr>-50*med_cor+50, "Good", "Bad")) %>%
  mutate(is_iso=ifelse(med_iso_cor>0.8, "Is iso", "Not iso")) %>%
  mutate(time_trend=ifelse(p_adj<0.001& estimate>0, "Increases", "No change"))
feat_shape_df_xamend %>%
  bind_rows(iso_df %>% mutate(amendment="Overall")) %>%
  filter(!is.na(time_trend)) %>%
  count(peak_qual, is_iso, time_trend, amendment) %>%
  pivot_wider(names_from = amendment, values_from=n, values_fill = 0) %>%
  print()
feat_shape_df_xamend %>%
  bind_rows(feat_shape_df %>% mutate(amendment="Overall")) %>%
  filter(!is.na(time_trend)) %>%
  ggplot() +
  geom_bar(aes(x=amendment)) +
  facet_wrap(~peak_qual+is_iso+time_trend, scales="free_y", ncol=4)
```

@@

```{r create a heatmap and cluster all high-quality peaks}
labeled_fracs <- cleaned_combined_peaks %>%
  summarise(med_iso_cor=median(iso_cor), 
            med_snr=median(beta_snr),
            med_cor=median(beta_cor),
            .by=c(feature)) %>%
  filter(med_snr>-50*med_cor+50) %>%
  filter(med_iso_cor>0.8) %>%
  select(feature) %>%
  left_join(cleaned_combined_peaks %>% select(feature, filename, lab_area, unlab_area), multiple = "all") %>%
  mutate(labeled_frac=lab_area/(lab_area+unlab_area)) %>%
  select(-lab_area, -unlab_area)

cmpd_order <- labeled_fracs %>%
  left_join(parametadata) %>%
  summarise(mean_label=mean(labeled_frac, na.rm=TRUE), .by=feature) %>%
  arrange(desc(mean_label)) %>%
  pull(feature)

colorpal <- viridis::viridis(10, alpha=seq(0, 1, length.out=10), direction = -1)
untarg_heatmap <- labeled_fracs %>%
  left_join(parametadata) %>%
  filter(samp_type=="Smp") %>%
  mutate(feature=factor(feature, levels=cmpd_order)) %>%
  ggplot() +
  geom_raster(aes(x=timepoint, y=tripl, fill=labeled_frac)) +
  scale_fill_gradientn(colours = colorpal, limits = c(0, 1)) +
  facet_wrap(feature~amendment+depth+startime) +
  scale_y_discrete(expand=expansion()) +
  scale_x_discrete(guide = guide_axis(n.dodge = 5), expand=expansion()) +
  facet_nested(feature~amendment+depth+startime, scales = "free_x", switch="y", strip=strip_nested(size = "variable")) +
  labs(x="Timepoint", y="Metabolite", fill="Fraction\nlabeled") +
  theme_bw() +
  theme(panel.spacing.x=unit(0.2, "lines"), 
        panel.spacing.y = unit(0, "lines"),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        # axis.ticks.length.x=unit(1:5, "lines"),
        # axis.text.x = element_markdown(vjust = -5:-1),
        strip.text.y.left = element_text(angle=0, hjust = 1),
        # strip.background.y = element_rect(fill=NA, color=NA),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "top", 
        legend.key.width = unit(1, "null")) +
  guides(
    fill=guide_colorbar(theme=theme(
      legend.frame = element_rect(color = "black"),
      legend.ticks = element_line(color="black"),
      legend.text = ggtext::element_markdown(hjust=c(0, 0.5, 0.5, 0.5, 1)))
    )
  )
ggsave("untargeted/untarg_heatmap.pdf", width = 10, height=36, units = "in")
```

>>>>

```{r get counts per amendment instead of as a whole}
# Repeat analysis across amendments to get additional numbers
lmtest_xamend <- cleaned_combined_peaks %>%
  arrange(feature) %>%
  select(feature, filename, into) %>%
  mutate(samp_type=str_extract(filename, "Smp|Blk|Poo")) %>%
  filter(samp_type=="Smp") %>%
  mutate(timepoint=as.numeric(str_extract(filename, "(?<=T)\\d+"))) %>%
  mutate(timepoint=ifelse(timepoint==0, 0.3, timepoint)) %>%
  mutate(amendment=str_extract(filename, "Amm|GMP|Arg|Nit")) %>%
  mutate(depth=str_extract(filename, "Surf|Deep")) %>%
  nest(data=-c(feature, amendment)) %>%
  filter(sapply(data, nrow)>2) %>%
  mutate(lm_test=map(data, ~broom::tidy(lm(log10(.x$into)~log10(.x$timepoint))), .progress = TRUE)) %>%
  unnest(lm_test) %>%
  filter(term=="log10(.x$timepoint)") %>%
  mutate(p_adj=p.adjust(p.value, method = "fdr"))
feat_shape_df_xamend <- cleaned_combined_peaks %>%
  mutate(amendment=str_extract(filename, "Amm|GMP|Arg|Nit")) %>%
  summarise(mzmed=unique(mzmed),
            rtmed=unique(rtmed),
            med_iso_cor=median(iso_cor), 
            med_snr=median(beta_snr),
            med_cor=median(beta_cor),
            .by=c(feature, amendment)) %>%
  left_join(lmtest_xamend %>% select(feature, amendment, estimate, p_adj)) %>%
  mutate(peak_qual=ifelse(med_snr>-50*med_cor+50, "Good", "Bad")) %>%
  mutate(is_iso=ifelse(med_iso_cor>0.8, "Is iso", "Not iso")) %>%
  mutate(time_trend=ifelse(p_adj<0.001& estimate>0, "Increases", "No change"))
feat_shape_df_xamend %>%
  bind_rows(iso_df %>% mutate(amendment="Overall")) %>%
  # filter(!rtmed%between%(c(14, 14.5)*60)) %>%
  # filter(!rtmed<(2*60)) %>%
  filter(!is.na(time_trend)) %>%
  count(peak_qual, is_iso, time_trend, amendment) %>%
  pivot_wider(names_from = amendment, values_from=n, values_fill = 0) %>%
  print()
feat_shape_df_xamend %>%
  bind_rows(feat_shape_df %>% mutate(amendment="Overall")) %>%
  filter(!is.na(time_trend)) %>%
  ggplot() +
  geom_bar(aes(x=amendment)) +
  facet_wrap(~peak_qual+is_iso+time_trend, scales="free_y", ncol=4)
```

@@

```{r heatmap of all high-quality feats}
high_qual_feats <- cleaned_combined_peaks %>%
  filter(rtmed>120 & !rtmed%between%c(14*60, 14.5*60)) %>%
  left_join(parametadata %>% distinct(filename, samp_type)) %>%
  filter(samp_type=="Smp") %>%
  summarise(med_iso_cor=median(iso_cor), 
            med_snr=median(beta_snr),
            med_cor=median(beta_cor),
            .by=c(feature)) %>%
  filter(med_snr>-50*med_cor+50) %>%
  filter(med_iso_cor>0.8)
untarg_lab_fracs <- cleaned_combined_peaks %>%
  filter(feature%in%high_qual_feats$feature) %>%
  select(feature, filename, lab_area, unlab_area) %>%
  left_join(parametadata %>% distinct(filename, samp_type)) %>%
  filter(samp_type=="Smp") %>%
  slice_max(unlab_area, by=c(feature, filename), with_ties = FALSE) %>%
  mutate(labeled_frac=lab_area/(lab_area+unlab_area)) %>%
  left_join(parametadata %>% distinct(filename, shortname)) %>%
  select(feature, shortname, labeled_frac)

cmpd_order <- untarg_lab_fracs %>%
  summarise(mean_label=mean(labeled_frac, na.rm=TRUE), .by=feature) %>%
  arrange(desc(mean_label)) %>%
  pull(feature)

colorpal <- viridis::viridis(10, alpha=seq(0, 1, length.out=10), direction = -1)
untarg_heatmap <- untarg_lab_fracs %>%
  complete(feature, shortname, fill=list(labeled_frac=0)) %>%
  mutate(feature=factor(feature, levels=cmpd_order)) %>%
  left_join(parametadata %>% filter(IS_type=="noIS")) %>%
  ggplot() +
  geom_raster(aes(x=timepoint, y=tripl, fill=labeled_frac)) +
  scale_fill_gradientn(colours = colorpal, limits = c(0, 1)) +
  scale_y_discrete(expand=expansion()) +
  scale_x_discrete(guide = guide_axis(n.dodge = 5), expand=expansion()) +
  facet_nested(feature~amendment+depth+startime, scales = "free_x", switch="y", strip=strip_nested(size = "variable")) +
  labs(x="Timepoint", y="Metabolite", fill="Fraction\nlabeled") +
  theme_bw() +
  theme(panel.spacing.x=unit(0.2, "lines"), 
        panel.spacing.y = unit(0, "lines"),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        # axis.ticks.length.x=unit(1:5, "lines"),
        # axis.text.x = element_markdown(vjust = -5:-1),
        strip.text.y.left = element_text(angle=0, hjust = 1),
        # strip.background.y = element_rect(fill=NA, color=NA),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "top", 
        legend.key.width = unit(1, "null")) +
  guides(
    fill=guide_colorbar(theme=theme(
      legend.frame = element_rect(color = "black"),
      legend.ticks = element_line(color="black"),
      legend.text = ggtext::element_markdown(hjust=c(0, 0.5, 0.5, 0.5, 1)))
    )
  )
ggsave("untargeted/untarg_heatmap.pdf", width = 10, height=36, units = "in")
```

```{r run kmeans}
untarg_lab_mat <- untarg_lab_fracs %>%
  bind_rows(labeled_fracs %>% rename(feature=compound_name)) %>%
  pivot_wider(names_from = "feature", values_from = "labeled_frac", values_fill=0) %>%
  column_to_rownames("shortname") %>%
  data.matrix()

set.seed(124)

# sapply(1:20, function(n_clust){
#   replicate(10, {
#     kmeans_out <- kmeans(t(untarg_lab_mat), centers = n_clust)
#     kmeans_out$betweenss / kmeans_out$totss * 100
#   })
# }) %>%
#   boxplot()

n_clust <- 10
untarg_kmeans <- untarg_lab_mat %>%
  t() %>%
  kmeans(centers = n_clust)
kmeans_df <- data.frame(cluster=1:n_clust, n=untarg_kmeans$size) %>%
  arrange(desc(n)) %>%
  mutate(cluster_name=paste0("Cluster ", row_number(), " n=", n)) %>%
  mutate(cluster_name=fct_inorder(cluster_name)) %>%
  left_join(data.frame(cluster=untarg_kmeans$cluster, feature=names(untarg_kmeans$cluster))) %>%
  select(cluster_name, feature)

# Plot clusters as heatmap
untarg_lab_fracs %>%
  left_join(kmeans_df) %>%
  summarise(labeled_frac=mean(labeled_frac), .by = c(cluster_name, shortname)) %>%
  left_join(parametadata) %>%
  ggplot() +
  geom_raster(aes(x=timepoint, y=tripl, fill=labeled_frac)) +
  scale_fill_gradientn(colours = colorpal, limits = c(0, 1)) +
  facet_wrap(cluster_name~amendment+depth+startime) +
  scale_y_discrete(expand=expansion()) +
  scale_x_discrete(guide = guide_axis(n.dodge = 5), expand=expansion()) +
  facet_nested(cluster_name~amendment+depth+startime, scales = "free_x", switch="y", 
               strip=strip_nested(size = "variable")) +
  labs(x="Timepoint", y=NULL, fill="Fraction\nlabeled") +
  theme_bw() +
  theme(panel.spacing.x=unit(0.2, "lines"), 
        panel.spacing.y = unit(0, "lines"),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        # axis.ticks.length.x=unit(1:5, "lines"),
        # axis.text.x = element_markdown(vjust = -5:-1),
        strip.text.y.left = element_text(angle=0, hjust = 1),
        # strip.background.y = element_rect(fill=NA, color=NA),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "top", 
        legend.key.width = unit(1, "null")) +
  guides(
    fill=guide_colorbar(theme=theme(
      legend.frame = element_rect(color = "black"),
      legend.ticks = element_line(color="black"),
      legend.text = ggtext::element_markdown(hjust=c(0, 0.5, 0.5, 0.5, 1)))
    )
  )

# Export as CSV
kmeans_df %>%
  filter(!startsWith(feature, "FT")) %>%
  arrange(cluster_name, feature) %>%
  split(.$cluster_name) %>%
  map(function(df_i){
    df_i %>% add_row(cluster_name="", feature="")
  }) %>%
  bind_rows() %>%
  write_csv("untargeted/cluster_assignments.csv")
```
